Type-Directed, Whitespace-Delimited Parsing for Embedded DSLs

### 2014-03-15 esop14/att-esop14.tex
removed stuff

\section{Motivation}
When designing and implementing a new abstraction, experts typically begin by attempting to define new constructs in terms of existing language constructs.
This approach is often effective because modern {general-purpose} abstraction mechanisms, like inductive datatypes and object systems, are highly expressive.  
For example, the Delite framework leverages Scala's powerful general-purpose mechanisms to enable a number of useful \emph{embedded domain-specific languages}.
%\cite{delite}
Unfortunately, there remain some situations of interest where general-purpose abstractions fall short. 
For instance, it is difficult to adequately encode advanced type systems in terms of the simpler rules governing general-purpose abstractions (e.g. reasoning about units of measure requires built-in language support in F\# \cite{conf/cefp/Kennedy09}). 
Even if a full encoding is possible, it may not be useful if it is overly verbose or unnatural, or if the error messages are overly abstract. For example, regular expressions encoded using inductive datatypes are often considered overly verbose, so most functional languages support them via strings, which is less safe. 
Finally, general-purpose abstractions are implemented in a uniform manner. Domain knowledge is not easily applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, particularly when targeting heterogeneous hardware platforms (e.g. programmable GPUs) and distributed computing resources. Using variants of familiar abstractions backed by specialized implementation strategies can be quite useful in these cases.

@@

Researchers or domain experts who run into situations like these, where more direct control over a language's semantics and implementation are needed, have little choice today but to realize new abstractions by creating a new language of some form. They might develop a new standalone language from scratch, modify an implementation of an existing language, or use tools like compiler generators, DSL frameworks and language workbenches \cite{fowler2010domain}. 
%In our simple scenario, we may simply fork our implementation of Godel's T or even edit it directly (a pernicious technique for implementing a new language where the prior one is overwritten). 
%In a more complex scenario, we may instead employ a tool like a compiler generator or DSL framework \cite{fowler2010domain} that can generate a standalone implementation from declarative specifications of language constructs. Some of these tools allow you to package and reuse these specifications (with the important caveat that not all combinations of constructs are valid and free of conflicts, an important modularity issue that we will return to several times in this paper).
The increasing sophistication and ease-of-use of these tools have led to calls for a {\it language-oriented approach} to software development, where different components of an application are written in different specialized languages \cite{journals/stp/Ward94}. Indeed, a number of software ecosystems are now explicitly designed to support many different languages, both general-purpose and domain-specific, atop a common intermediate language. The Java virtual machine (JVM), the Common Language Infrastructure (CLI) and LLVM are prominent examples of such ecosystems.

Unfortunately, this leads to a critical problem at language boundaries: a library's external interface must only use constructs that can reasonably be expressed in \emph{all possible client languages}. This discourages languages from including constructs that rely on statically-checked invariants stronger than those supported by their underlying implementation in the common intermediate language. At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee these invariants. This compromises both verifiability and performance. For%requires the development of an interoperability layer for every pair of DSLs. 
Moreover, this approach exposes the internals of an implementation to clients, making the abstraction awkward to work with and causing code breakage when implementation details change. This defeats a primary purpose of high-level programming languages: hiding low-level details from clients of an abstraction. We diagram this fundamental \emph{interoperability problem} in Figure \ref{approaches}(a). 
%As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be included in the implementation.
\begin{figure}[t]
\vspace{-15px}
\includegraphics[scale=0.415]{approaches.pdf}
\vspace{-30px}
\caption{(a) With the language-oriented approach, different primitive abstractions are packaged into separate languages that extend and target a common intermediate language (e.g. JVM bytecode). Users can only interface with libraries written in another language via the constructs in the common language, causing \emph{interoperability problems}. (b) With the language-internal approach, the semantics of new abstractions (i.e. the logic that governs typechecking and translation to a fixed {internal language}, here labeled \texttt{I}) can be implemented directly within so-called \emph{active libraries}. Clients can import and use these abstractions directly whenever needed.}
\vspace{-10px}
\label{approaches}
\end{figure}

>>>>

\section{Motivation}
When designing and implementing a new abstraction, experts typically begin by attempting to define new constructs in terms of existing language constructs.
This approach is often effective because modern {general-purpose} abstraction mechanisms, like inductive datatypes and object systems, are highly expressive.  
For example, the Delite framework leverages Scala's powerful general-purpose mechanisms to enable a number of useful \emph{embedded domain-specific languages} \cite{delite}.
Unfortunately, there remain some situations of interest where general-purpose abstractions fall short. 
For instance, it is difficult to adequately encode advanced type systems in terms of the simpler rules that govern general-purpose abstractions (e.g. reasoning about units of measure requires built-in language support in F\# \cite{conf/cefp/Kennedy09}). 
Even if a full encoding is possible, it may not be useful if it is overly verbose or unnatural, or if the error messages are overly abstract. For example, regular expressions encoded using inductive datatypes are quite verbose, so most functional languages support them via strings, which is less safe. 
Finally, general-purpose abstractions are implemented in a uniform manner, meaning domain-specific heuristics cannot be applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, such as when targeting heterogeneous hardware platforms (e.g. programmable GPUs) and distributed computing resources.

@@

Researchers or domain experts who run into situations like these, where more direct control over a language's semantics and implementation are needed, have little choice today but to realize new abstractions by creating a new language in some way. They might develop a new standalone language from scratch, modify an implementation of an existing language, or use tools like compiler generators, DSL frameworks and language workbenches \cite{fowler2010domain}. 
%In our simple scenario, we may simply fork our implementation of Godel's T or even edit it directly (a pernicious technique for implementing a new language where the prior one is overwritten). 
%In a more complex scenario, we may instead employ a tool like a compiler generator or DSL framework \cite{fowler2010domain} that can generate a standalone implementation from declarative specifications of language constructs. Some of these tools allow you to package and reuse these specifications (with the important caveat that not all combinations of constructs are valid and free of conflicts, an important modularity issue that we will return to several times in this paper).
The increasing sophistication and ease-of-use of these tools have led to calls for a {\it language-oriented approach} to software development, where different components of an application are written in different specialized languages \cite{journals/stp/Ward94}. Indeed, a number of software ecosystems are now designed explicitly to support many different languages, both general-purpose and domain-specific, atop a common intermediate language. The Java virtual machine (JVM), the Common Language Infrastructure (CLI) and LLVM are prominent examples of such ecosystems.

Unfortunately, this leads to a critical problem at language boundaries: a library's external interface must only use constructs that can reasonably be expressed in \emph{all possible client languages}. This discourages languages from including constructs that rely on statically-checked invariants stronger than those supported by their underlying implementation in the common intermediate language. At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee these invariants. This compromises both verifiability and performance. %Forrequires the development of an interoperability layer for every pair of DSLs. 
Moreover, this approach exposes the internals of an implementation to clients, making the abstraction awkward to work with and causing code breakage when implementation details change. This defeats a primary purpose of high-level programming languages: hiding low-level details from clients of an abstraction. We diagram this fundamental \emph{interoperability problem} in Figure \ref{approaches}(a). 
%As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be included in the implementation.
\begin{figure}[t]
\vspace{-15px}
\includegraphics[scale=0.415]{approaches.pdf}
\vspace{-30px}
\caption{(a) With the language-oriented approach, different primitive abstractions are packaged into separate languages that extend and target a common intermediate language (e.g. JVM bytecode). Users can only interface with libraries written in another language via the constructs in the common language, causing \emph{interoperability problems}. (b) With the language-internal approach, the semantics of new abstractions (i.e. the logic that governs typechecking and translation to a fixed {internal language}, here labeled \texttt{I}) can be implemented directly within so-called \emph{active libraries}. Clients can import and use these abstractions directly whenever needed.}
\vspace{-10px}
\label{approaches}
\end{figure}

### 2014-03-15 globaldsl13/motivation.tex
removed stuff

Figure~\ref{f:dsl-query} shows how a DSL for database queries can be used from within ordinary Wyvern code.  The example shows code for computing a feed that is derived from tweets about a company's new products.  In this example, the use of a querying DSL is triggered by the use of methods named \lstinline{query} expecting an argument of type \lstinline{DBQuery} (line 1) or \lstinline{FeedQuery} (line 5) respectively.  These types define related but distinct syntax for queries, determined by the expected type of expression where the tilde appears (tildes need not appear only at the ends of lines). Queries are again delimited by indentation. This mechanism is similar to what can be expressed in languages with built-in query syntax like LINQ \cite{mslinq}, but in this case, it is entirely user-defined, rather than built into the language.\begin{figure}
  \centering
  \begin{lstlisting}
serve(page, loc) where 
  val page = ~ 
    html:
      head:
        title: Hot Products
        style: {myStylesheet}
        body:
          div id="search":
            {SearchBox("products")}
          div id="products":
            {FeedBox(servlet.hotProds())}
  val loc = ~
    products.nameless.com
  \end{lstlisting}
  \caption{Wyvern DSLs: Presentation and URLs}
  \label{f:dsl-presentation}
\end{figure}

>>>>

Figure~\ref{f:dsl-query} shows how a DSL for database queries can be used from within ordinary Wyvern code.  The example shows code for computing a feed that is derived from tweets about a company's new products.  In this example, the use of a querying DSL is triggered by the use of methods named \lstinline{query} expecting an argument of type \lstinline{DBQuery} (line 1) or \lstinline{FeedQuery} (line 5) respectively.  These types define related but distinct syntax for queries, determined by the expected type of expression where the tilde appears (tildes need not appear only at the ends of lines). Queries are again delimited by indentation. This mechanism is similar to what can be expressed in languages with built-in query syntax like LINQ \cite{mslinq}, but in this case, it is entirely user-defined, rather than built into the language.\begin{figure}
  \centering
  \begin{lstlisting}
serve : (HTML, URL) -> unit
serve(~, `products.nameless.com`) 
  html:
    head:
      title: Hot Products
      style: {myStylesheet}
      body:
        div id="search":
          {SearchBox("products")}
        div id="products":
          {FeedBox(servlet.hotProds())}
  \end{lstlisting}
  \caption{Wyvern DSLs: Presentation and URLs}
  \label{f:dsl-presentation}
\end{figure}
